<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{Bootstrapping for Propensity Score Analysis}
-->

# Bootstrapping for Propensity Score Analysis

## Jason M. Bryer [jason@bryer.org](mailto:jason@bryer.org)
#### Project website: [jason.bryer.org/PSAboot](http://jason.bryer.org/PSAboot)
#### Bug Reports: [github.com/jbryer/PSAboot](https://github.com/jbryer/PSAboot)

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
require(knitr)
require(reshape)
opts_chunk$set(comment='')
require(PSAboot)
```

# Abstract

As the popularity of propensity score methods for estimating causal effects in observational studies increase, the choice of specific method has also increased. Rosenbaum (2012) has suggested that there are substantial benefits for testing a hypothesis more than once. Moreover, with the wide availability of high power computers resampling methods such as bootstrapping (Efron, 1979) have become popular for providing more stable estimates of the sampling distribution. This paper introduces the `PSAboot` package for R that provides functions for bootstrapping propensity score methods. It deviates from traditional bootstrapping methods by allowing for different sampling specifications for treatment and control groups. Additionally, this framework will provide estimates using multiple methods for each bootsrap sample. Two examples are discussed: the classic National Work Demonstration and PSID (Lalonde, 1986) study and a study on tutoring effects on student grades.

-----------------------------

# Introduction

\
The latest version of the `PSAboot` package can be downloaded from Github using the `devtools` package.

```{r, eval=FALSE}
devtools::install_github('PSAboot','jbryer')
```


The `PSAboot` function will perform the actual bootstrapping. It has a number of parameters for to specify how the bootsrap samples are drawn.

* `Tr` - a numeric (0 for control and 1 for treatment) or logical vector of treatment indicators.
* `Y` - a numeric vector with the outcome of interest.
* `X` - a data frame of covaraites.
* `M` (default is 100) - the number of bootstrap samples to draw.
* `formu` - the formula for estimating the propensity scores in phase I. Note that the dependent variable does not need to be specified as it will be replaced when combining the `Tr` vector and `X` data frame.
* `control.ratio` (default is 5) - This specifies the sample size of control units as a ratio of treatment units. For example, with a default value of 5 and 100 treatment units, this methods will sample 500 control units for each sample, or the number of control units if smaller than 500. When the ratio of treatment-to-control units increases, the range of propensity scores (using logistic regression) shrinks. Randomly selecting a subset of control units often results in wider and better overalapping distribution of propensity scores. See the the [`PSranges`](http://jason.bryer.org/multilevelPSA/psranges.html) function in the [`multilevelPSA`](http://jason.bryer.org/multilevelPSA) package for more information or this page: [http://jason.bryer.org/multilevelPSA/psranges.html](http://jason.bryer.org/multilevelPSA/psranges.html).
* `control.sample.size` (default is 5 times the number of treatment units) - The number of control units to sample for each bootstrap sample. If specified, this overrides the `control.ratio` parameter.
* `control.replace` (default is `TRUE`) - Specify whether random sampling of control units does so with replacement.
* `treated.sample.size` (default is the number of treatment units) - The number of treatment units to sample for each bootstrap sample.
* `treated.replace` (default is `TRUE`) - Specify whether random sampling of treatment units does so with replacement.
* `methods` - A list of functions to perform a propensity score analysis for each bootstrap sample. See the methods section below.
* `parallel` (default is `TRUE`) - Whether the bootstraping procedure should be run in parallel.
* `seed` - Seed used for the random number generator. If specified, the random seed will be set to `seed + i` where `i` is the current bootsrap sample in (1, M).

Other parameters can be passed to `methods` using the `...` parameter.

## Methods

The `methods` parameter on the `PSAboot` function specifies the different propensity score methods that will be used. Five methods are included, they are:

* `boot.strata` - This method estimates propensity scores using logistic regression and stratifies using quintiles on the propensity scores. Effects within each strata are estimated and aggregated.
* `boot.ctree` - This method creates strata using conditional inference trees vis-à-vis the `ctree` function in the `party` package. Effects within each strata (i.e. leaf node) are estimated and aggregated.
* `boot.rpart` - This method creates strata using classification trees vis-à-vis the `rpart` function. Effects within each strata (i.e. leaf node) are estimated and aggregated.
* `boot.matching` - This method finds matched pairs using the `Match` function in the `Matching` package. A paired dependent sample t-test is used to estimate effect sizes.
* `boot.matchit` - This method finds match pairs using the `matchit` function in the `MatchIt` package. A paired dependent sample t-ttest is used to estimate effect sizes.

### Defining Custom Methods

It is possible to define a custom method. Simply define a function with, at minimum, the following six parameters:

* `Tr` - A logicial or integer (0 and 1) vector with treatment indicators.
* `Y` - A numeric vector represeting the outcome.
* `X` - A data frame with the covariates.
* `X.trans` - A data frame with factor levels dummy coded.
* `formu` - A formula for estimating propensity scores in phase one.
* `...` - Other parameters passed by the user.

Each method must return a `list` with three elements:

* `summary` - This must be a named numeric vector with at minimum `estimate`, `ci.min`, and `ci.max`, however other values allowed.
* `balance` - This must be a named numeric vector with one element per covariate listed in `X.trans` representing a balance statistic. It is recommended, and the implementation for the built-in methods, to use an absolute standardized effect size. As will be shown below, the summary and plotting functions will include an unadjusted balance statistic (i.e. effect size) before adjustment for comparison.
* `details` - This can be an arbitrary object, typically the result of the underlying method used.

For example, the `boot.matching.1to3` function below wraps the built-in `boot.matching` method but sets the `M` parameter to 3, thereby performing 1-to-3 matching instead of the default 1-to-1 matching. This framework simplifies the process of using, and comparing, slight variations of different propensity score methods. 

```{r}
boot.matching.1to3 <- function(Tr, Y, X, X.trans, formu, ...) {
	return(boot.matching(Tr=Tr, Y=Y, X=X, X.trans=X.trans, formu=formu, M=3, ...))
}
```

The `PSAboot` function returns an object of class `PSAboot`. The following S3 methods are implemented: `print`, `summary`, `plot`, `boxplot`, and `matrixplot`.

-----------------------------

# Example One: National Work Demonstration and PSID

The `lalonde` (Lalonde, 1986) has become the *de defacto* teaching dataset in PSA since Dehejia and Wahba's (1999) re-examination of the National Supported Work Demonstration (NSW) and the Current Population Survey (CPS). 

The `lalonde` data set is included in the `MatchIt` package. The crosstab shows that there are 429 control units and 185 treatment units.

```{r lalonde.load}
data(lalonde, package='MatchIt')
table(lalonde$treat)
```
```{r lalonde.boot, cache=TRUE}
lalonde.formu <- treat ~ age + I(age^2) + educ + I(educ^2) + black +
	hispan + married + nodegree + re74  + I(re74^2) + re75 + I(re75^2) +
	re74 + re75

boot.lalonde <- PSAboot(Tr=lalonde$treat, Y=lalonde$re78,
						X=lalonde[,c('re74','re75','educ','black','hispan','age','married','nodegree')],
						M=100, seed=2112)
```

```{r}
summary(boot.lalonde)
```

```{r lalonde.plot}
plot(boot.lalonde)
```

```{r lalonde.histogram, warning=FALSE, message=FALSE}
hist(boot.lalonde)
```


```{r lalonde.boxplot}
boxplot(boot.lalonde)
```

```{r lalonde.matrixplot}
matrixplot(boot.lalonde)
```

## Evaluating Balance

```{r lalonde.balance, cache=TRUE}
lalonde.bal <- balance(boot.lalonde)
lalonde.bal
```

```{r lalonde.balance.plot}
plot(lalonde.bal)
```

```{r lalonde.balance.boxplot}
boxplot(lalonde.bal)
```

-----------------------------

# Example Two: Effectiveness of Tutoring on Course Grades

```{r tutoring.setup}
require(TriMatch)
require(PSAboot)
data(tutoring, package='TriMatch')
tutoring$treatbool <- tutoring$treat != 'Control'
covs <- tutoring[,c('Gender', 'Ethnicity', 'Military', 'ESL', 'EdMother', 'EdFather',
					'Age', 'Employment', 'Income', 'Transfer', 'GPA')]

table(tutoring$treatbool)
```

```{r tutoring.psaboot, cache=TRUE}
tutoring.boot <- PSAboot(Tr=tutoring$treatbool, 
						 Y=tutoring$Grade, 
						 X=covs, 
						 seed=2112,
						 control.sample.size=918, control.replace=TRUE,
						 treated.sample.size=224, treated.replace=TRUE,
						 methods=c('Stratification'=boot.strata,
						 		  'ctree'=boot.ctree,
						 		  'rpart'=boot.rpart,
						 		  'Matching'=boot.matching,
						 		  'Matching-1-to-3'=boot.matching.1to3,
						 		  'MatchIt'=boot.matchit)
)
summary(tutoring.boot)
```


```{r tutoring.plot}
plot(tutoring.boot)
```

```{r tutoring.histogram, message=FALSE, warning=FALSE}
hist(tutoring.boot)
```

```{r tutoring.boxplot}
boxplot(tutoring.boot)
```

```{r tutoring.matrixplot}
matrixplot(tutoring.boot)
```

```{r tutroing.balance}
tutoring.bal <- balance(tutoring.boot)
tutoring.bal
plot(tutoring.bal)
boxplot(tutoring.bal)
```

-----------------------------

# References

Bryer, J.M. (2013). TriMatch: Propensity Score Matching of Non-Binary Treatments. R package version 0.9.1. http://jason.bryer.org/TriMatch

Helmreich, J.E., Pruzek, R.M. (2009). PSAgraphics: An R Package to Support Propensity Score Analysis. Journal of Statistical Software 29(6), 1-23. http://www.jstatsoft.org/v29/i06/.

Ho, D.E., Imai, K., King, G., & Stuart, E.A. (2011). MatchIt: Nonparametric Preprocessing for Parametric Causal Inference. Journal of Statistical Software, (42) 8, 1-28. URL http://www.jstatsoft.org/v42/i08/

Hothorn, T., Hornik, K., & Zeileis, A. (2006). Unbiased Recursive Partitioning: A Conditional Inference Framework. Journal of Computational and Graphical Statistics, 15(3), 651-674.

Lalonde, R. (1986). Evaluating the econometric evaluations of training programs with experimental data. American Economic Review 76: 604-620.

Dehejia, R.H. and Wahba, S. (1999). Causal Effects in Nonexperimental Studies: Re-Evaluating the Evaluation of Training Programs. Journal of the American Statistical Association 94: 1053-1062.

R Core Team (2013). R: A language and environment for statistical computing. R Foundation for Statistical Computing, Vienna, Austria. URL http://www.R-project.org/.

Rosenbaum, P.R. (2012). Testing one hypothesis twice in observational studies. Biometrika, 99, 4, 763-774.

Rosenbaum, P.R., & Rubin, D.B. (1983). The central role of the propensity score in observational studies for causal effects. Biometrika, 70, 1, 41-55.

Sekhon, J.S. (2011). Multivariate and Propensity Score Matching Software with Automated Balance Optimization: The Matching Package for R. Journal of Statistical Software, 42(7), 1-52. URL http://www.jstatsoft.org/v42/i07/.

Therneau, T., Atkinson, B., & Ripley, B. (2013). rpart: Recursive Partitioning. R package version 4.1-3. http://CRAN.R-project.org/package=rpart

